\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
More bits
Add more delectable references and some text.

With the advent of high-throughput coprocessors, such as GPGPUs or the MIC
chip, comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
subcycling).  When using subcycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.

@@

For hyperbolic problems, we integrate the solution on a a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non- linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.

>>>>

With the advent of many-core chips, such as GPUs and the MIC architecture, comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin via methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense computational patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
% dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  When using sub-cycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  Additionally many-core implementations, such as \manyclaw \cite{manyclaw}, can be used for the hyperbolic integration on a leaf-patch.  The design of leaf-patches allows for the smart management of data local to accelerator and the host.

@@

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non- linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.
